---
title: "ITScontrol_demonstration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ITScontrol_demonstration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(multipleITScontrol)
library(pheicharts)
library(tidyverse)


```

## Usage

This is a basic example which shows you how to solve a common problem with two stage interrupted time series with a control:

**Background**: Alpine Meadow School and Forest Tiger School have similar student demographics, including socioeconomic status, ethnicity, and academic performance. Both schools are part of Clarkson County's public school district.

Alpine Meadow School wants to trial out two new interventions to improve their school's reading comprehension score, and to compare post intervention results with the pre-intervention score.

**Intervention 1: Implementing a New Reading Programme**

-   **Objective:** Improve reading comprehension and literacy rates among students.
-   **Start Date:** September 1, 2025
-   **Duration:** 6 months
-   **Description:** The school introduces a new, evidence-based reading program that includes daily reading sessions, interactive reading activities, and regular assessments.
-   **Measurement:** Reading comprehension scores from standardized tests administered weekly.

**Intervention 2: Introducing Peer Tutoring Sessions**

-   **Objective:** Further enhance reading comprehension and literacy rates.
-   **Start Date:** March 1, 2026 (immediately after the reading program ends)
-   **Duration:** 6 months
-   **Description:** The school implements peer tutoring sessions where older students tutor younger students in reading. These sessions are held twice a week and focus on reading comprehension strategies and practice.
-   **Measurement:** Reading comprehension scores from standardized tests administered weekly.

### Controlled Interrupted Time Series Design

**Step 1: Baseline Period**

-   **Duration:** 6 months (March 1, 2025 - August 31, 2025)
-   **Data Collection:** Collect baseline data on reading comprehension scores administered weekly.

**Step 2: Intervention 1 Period**

-   **Duration:** 6 months (September 1, 2025 - February 28, 2026)
-   **Data Collection:** Continue collecting data on reading comprehension scores during the reading program administered weekly.

**Step 3: Intervention 2 Period**

-   **Duration:** 6 months (March 1, 2026 - August 31, 2026)
-   **Data Collection:** Collect data on reading comprehension scores during the peer tutoring sessions administered weekly.

The calendar plot below summarises the timeline of the interventions:

```{r calendar, echo = FALSE, warning = FALSE, message = FALSE, fig.height=10, fig.retina=3}

tibble_data <- tibble::tibble(Date = seq(as.Date("2025-03-03"), as.Date("2026-08-30"), by = "day"),
  Period = dplyr::case_when(
    Date >= as.Date("2025-03-03") & Date <= as.Date("2025-08-31") ~ "Pre-intervention period",
    Date >= as.Date("2025-09-01") & Date <= as.Date("2026-03-01") ~ "Intervention 1) Reading Program",
    Date >= as.Date("2026-03-02") & Date <= as.Date("2026-08-30") ~ "Intervention 2) Peer Tutoring Sessions"
  ))

plot <- phei$phei_calendar(
  tibble_data,
  date_column = "Date",
  "Period",
  colours = phei_palettes$cool_colours,
  ncol = 3
) +
  theme(strip.text = element_text(size = rel(0.5)),
        axis.text = element_text(size = rel(0.5)),
        plot.caption = element_text(size = rel(0.5)),
        legend.text = element_text(size = rel(0.5)))
  

plot$layers[[2]]$aes_params$size <- 1.5

plot


```

# Step 1) Processing the Data

```{r example}
library(multipleITScontrol)



```
